Schema Change: Challenge Problems

### 2024-04-29 paper.tex
Revising the conference organizer section

\usepackage{multicol}
\usepackage{ccicons}
\usepackage{natbib}

@@

TODO: Need to explain that we take the programming system view - program is not something
built centrally, but something that exists and can be modified

@@

\section{Conference Organizer: Merging structural document edits}

\subsection{Context}
In this challenge, we consider a document-based format akin to that used, for example, by Notion. We focus on schema change applied to data, but the document-based format can also be extended to include programming as done, for example in Boxer. We assume that the document can be concurrently edited by multiple users in the same fashion as in local-first software, i.e., users may edit a local copy that then needs to be synchronized with changes made by other users. The edits to the document can be done in a rich-text editor, or by using a more structure-aware structure editor that can, for example, record changes as a sequence of high-level edit operations.

\subsection{Example}
In this challenge, we consider the problem of collaborative conference planning. A number of co-organizers are planning a conference that will feature talks by several invited speakers. The organizers need to agree on a list of speakers to invite, coordinate who contacts whom and track the acceptance of invitations. They start with a document shown in Figure~\ref{fig:conf-orig}.

\begin{figure}[h!]
\fbox{\parbox{24em}{
\textbf{PROGRAMMING CONFERENCE 2023}\\
\textbf{Invited speakers}
\begin{itemize}
 \item Adele Goldberg, adele@xerox.com
 \item Margaret Hamilton, hamilton@mit.com
 \item Betty Jean Jennings, betty@rand.com
\end{itemize}
}}
\caption{Conference organization. Initial state of the document.}
\label{fig:conf-orig}
\end{figure}

\begin{figure}
\fbox{\parbox{24em}{
\textbf{PROGRAMMING CONFERENCE 2023}\\
\textbf{Invited speakers}
\begin{itemize}
 \item Ada Lovelace, lovelace@rsoc.ac.uk
 \item Adele Goldberg, adele@xerox.com
 \item Betty Jean Jennings, betty@rand.com
 \item Margaret Hamilton, hamilton@mit.com
\end{itemize}
}}
\caption{Conference organization. Added a speaker and sorted list.}
\label{fig:conf-add}
\end{figure}

\begin{figure}
\fbox{\parbox{24em}{
\textbf{PROGRAMMING CONFERENCE 2023}\\
\textbf{Invited speakers}

\begin{tabular}{| l | l | l |}
  \hline
  \textbf{Name} & \textbf{Email} & \textbf{Organizer} \\
  \hline \hline
  Adele Goldberg & adele@xerox.com & TP\\
  \hline
  Margaret Hamilton & hamilton@mit.com & JE\\
  \hline
  Betty Jean Jennings & betty@rand.com & JE\\
  \hline
\end{tabular}
}}\caption{Conference organization. Refactored list into a table.}
\label{fig:conf-tab}
\end{figure}

\begin{figure}
\fbox{\parbox{24em}{
\textbf{PROGRAMMING CONFERENCE 2023}\\
\textbf{Invited speakers}

\begin{tabular}{| l | l | l |}
  \hline
  \textbf{Name} & \textbf{Email} & \textbf{Organizer} \\
  \hline \hline
  Ada Lovelace & lovelace@rsoc.ac.uk & \\
  \hline
  Adele Goldberg & adele@xerox.com & TP\\
  \hline
  Betty Jean Jennings & betty@rand.com & JE\\
  \hline
  Margaret Hamilton & hamilton@mit.com & JE\\
  \hline
\end{tabular}
}}\caption{Conference organization. Resulting document after change merging.}
\label{fig:conf-fin}
\end{figure}

\subsection{Challenge}
The challenge is to merge document edits that are done locally and independently by the co-organizers of the conference. Specifically, consider the following two edits:

@@

The system should be able to merge the changes and produce a final table shown in Figure~\ref{fig:conf-fin}. The resulting table needs to include the additional speaker created by the first organizer, use the order specified by the first organizer and use the format defined by the second organizer. The newly added speaker should be reformatted into the new format. For the newly added ``Organizer'' column, the second user may specify default value or the newly added row may use an empty value.

\subsection{Goals}
A solution to the challenge should strive to satisfy the following goals:

\begin{itemize}
\item \emph{Completeness.} It should be possible to semi-automatically merge any two sequences of edits. In other words, the merging should always be defined, regardless of what edits the users perform.
\item \emph{Commutativity.} The order of edits should not matter. If users $A$ and $B$ perform edits independently starting from the same initial document, the resulting document should be the same if the system treats edits from $A$ as occurring before the edits of $B$ and vice versa.
\item \emph{Conflict resolution.} It may not always be possible to merge conflicting edits. In this case, the system should interactively ask the user for guidance or, possibly, use default behavior specified by the user.
\item \emph{Convergence.} The system does not need to consider the case where one users does not want to adopt changes created by another user. We assume that all users want to eventually see the same final document produced by applying all edits.
\item \emph{Structure editing.} The system may use a structure editor that provides high-level commands for operations such as reordering of elements or refactoring of a list into a table. In other words, we assume that the challenge may not be solvable in a system based on plain text editing of document source code.
\end{itemize}

\subsection{Remarks}
It is worth noting that this challenge does not discuss explicit distinction between the schema (or ``type'') of the document and data (or ``values''). The implementing system may or may not maintain such distinction. In our example, adding a new speaker and sorting the list is arguably a mere change of the data, but the refactoring of a list into a table is a schema change. The system may leverage knowledge about such distinction and, for example, handle the merging of schema-related and data-related changes differently.

\section{Conference Organizer: Merging edits in the presence of code}

\subsection{Context}
Consider the document-based format as in the previous challenge, but extended with the support for computed values. A computed value is a node in the document whose value is specified by a formula akin to those used in spreadsheets. The equations can refer to other nodes in the document, possibly using either an absolute or relative path and may involve a range of built-in functions, for example to count the number of elements or to sum a sequence of numerical values. We now also assume that elements in the document can have IDs that are not displayed in the default view, but may be used to refer to them in equations.

\begin{figure}
\fbox{\parbox{28em}{
\textbf{PROGRAMMING CONFERENCE 2023}\\
\textbf{Invited speakers}
\begin{itemize}
 \item Adele Goldberg, adele@xerox.com
 \item Margaret Hamilton, hamilton@mit.com
 \item Betty Jean Jennings, betty@rand.com
\end{itemize}
\vspace{0.5em}
\textbf{Conference budget}\\
Travel cost per speaker:\\
\hspace*{2em} \$1200 \\
Number of speakers: \\
\hspace*{2em} \texttt{=COUNT(/ul[id='speakers']/li)} \\
Travel expenses:\\
\hspace*{2em} \texttt{=/dl/dd[0] * /dl/dd[1]}
}}
\caption{Conference organization. Initial state of the document.}
\label{fig:conf-budget}
\end{figure}

\begin{figure}
\fbox{\parbox{28em}{
\textbf{PROGRAMMING CONFERENCE 2023}\\
\textbf{Invited speakers}

\begin{tabular}{| l | l | l |}
  \hline
  \textbf{Name} & \textbf{Email} & \textbf{Organizer} \\
  \hline \hline
  Ada Lovelace & lovelace@rsoc.ac.uk & \\
  \hline
  Adele Goldberg & adele@xerox.com & TP\\
  \hline
  Betty Jean Jennings & betty@rand.com & JE\\
  \hline
  Margaret Hamilton & hamilton@mit.com & JE\\
  \hline
\end{tabular}

\vspace{0.5em}
\textbf{Conference budget}\\
Travel cost per speaker:\\
\hspace*{2em} \$1200 \\
Number of speakers: \\
\hspace*{2em} \texttt{=COUNT(/table[id='speakers']/tbody/tr)} \\
Travel expenses:\\
\hspace*{2em} \texttt{=/dl/dd[0] * /dl/dd[1]}
}}\caption{Conference organization. Resulting document after change merging.}
\label{fig:conf-finfin}
\end{figure}

\subsection{Example}
This challenge extends the previous problem. The conference organizers also want to use the document to manage the conference budget. In a very simplified form, one aspect of this is calculating the estimated travel expenses for all the speakers. For this, they will want to add a computed value to the document that counts the number of speakers and multiplies the result by a fixed cost per speaker elsewhere in the document.

The example is shown in Figure~\ref{fig:conf-budget}. The additional section of the document contains one fixed value (travel cost per speaker) and two computed values. The first selects all \texttt{li} elements of a \texttt{ul} element with ID \texttt{speakers} and counts their number. The second computed value then multiplies the constant travel cost per speaker by the number of speakers. Here, we assume the new section uses the HTML definition list \texttt{dl} and the equation selects its first and second \texttt{dd} item, respectively.

\subsection{Challenge}
The challenge is, again, to merge the edits done independently by the co-organizers. Assume two co-organizers start with the same original document shown in Figure~\ref{fig:conf-orig}.
\begin{enumerate}
    \item The first co-organizer performs the edits outlined previously that result in the document shown in Figure~\ref{fig:conf-fin}, i.e., edits the speakers and refactors the structure to use a table.
    \item The second co-organizer adds the budget calculation as shown in Figure~\ref{fig:conf-budget}.
\end{enumerate}
As before, the system should be able to merge the edits made to the document. Crucially, this requires that the refactoring of the document structure done in (1), is also reflected in the code of the equations added in (2). As shown in Figure~\ref{fig:conf-finfin}, in this specific case, the system changes the equation \texttt{COUNT(/ul[id='speakers']/li)} to \texttt{COUNT(/table[id='speakers']/tbody/tr)}. This is the case because the refactoring consists of four steps that each transform the document and also need to update the code of the equation accordingly:
\begin{itemize}
    \item[--] Change the type of the \texttt{ul} element to \texttt{table}
    \item[--] Wrap the body of the \texttt{table} element with \texttt{tbody}
    \item[--] Change the type of all children of \texttt{tbody} from \texttt{li} to \texttt{tr}
    \item[--] Further split the body into two \texttt{td} elements
\end{itemize}

\subsection{Goals}
A solution to the challenge should meet the goals of the preceding challenge, i.e., completeness (merging should be always defined), commutativity (order does not matter), conflict resolution (ask for guidance if needed), convergence (everyone eventually wants the same document) and structure editing (the system may capture high-level edits).

\subsection{Extensions}
The basic challenge outlined above focuses primarily on the local-first software scenario, but it can be extended to also apply to the live programming scenario. In particular, when the document contains computed values, those may be evaluated either explicitly by the user or automatically on-the-fly. An edit can then invalidate some of the values (either by changing the data that a computed value depends on, or by changing the equation used to compute the value). A live programming system should be able to detect which computed values are affected by an edit and, either invalidate those (requiring an explicit re-evaluation by the user) or automatically re-evaluate them (and possibly highlight the affected values to inform the user).

>>>>

\usepackage{changepage}
\usepackage{multicol}
\usepackage{ccicons}
\usepackage{natbib}
\usepackage[many]{tcolorbox}

\newcounter{challengen}
\newcommand{\challenge}[1]{\subsection*{\addtocounter{challengen}{1} Challenge \#\thechallengen: #1}}

\DeclareRobustCommand{\frameworkbox}[2][gray!15]{
\begin{tcolorbox}[breakable,left=3pt,right=3pt,top=3pt,bottom=3pt,colback=#1,colframe=#1,parbox=false,
  width=\dimexpr\textwidth\relax,enlarge left by=0mm,boxsep=5pt,arc=0pt,enlarge top by=0.5em,%enlarge bottom by=0.0em,
  outer arc=0pt]\setlength{\parskip}{0.5em}\setlength{\parindent}{0em}\textbf{\sffamily Framework perspective.}\quad #2
\end{tcolorbox}}

@@

\textcolor{red}{TODO: Need to explain that we take the programming system view - program is not something
built centrally, but something that exists and can be modified}

\textcolor{red}{TODO: Maybe the right way to frame this is to say that the challenges are intended
less for evaluating particular solutions but more for framing and making sense of different
kinds of schema-chaange-related problems.  You can read this paper and when you face a schema-change-related
problem, think about it in terms of our dimensions. Maybe compare it to our challenges.
We do not necessarilly expect that people will directly solve our challenges though. (Although
some might and that would be nice..)}

\textcolor{red}{TODO: I also guess this is more about just ``change'' than ``schema change''
which is one particular (most challenging) instance - i.e., change at the most permanent
pace layer.}

@@

\begin{figure}
\begin{adjustwidth}{-5.5em}{-5.5em}
\centering
\begin{subfigure}[b]{23em}\vspace{0pt}
   \fbox{\parbox{22.6em}{
   \textbf{PROGRAMMING CONFERENCE 2023}\\
   \textbf{Invited speakers}
   \begin{itemize}
    \item Adele Goldberg, adele@xerox.com
    \item Margaret Hamilton, hamilton@mit.com
    \item Betty Jean Jennings, betty@rand.com
   \end{itemize}
   \\~
   }}
   \caption{Initial state of the document.}
   \label{fig:conf-orig}
\end{subfigure}
\hfill
\begin{subfigure}[b]{23em}\vspace{0pt}
   \fbox{\parbox{22.6em}{
   \textbf{PROGRAMMING CONFERENCE 2023}\\
   \textbf{Invited speakers}
   \begin{itemize}
    \item Ada Lovelace, ada@rsoc.ac.uk
    \item Adele Goldberg, adele@xerox.com
    \item Betty Jean Jennings, betty@rand.com
    \item Margaret Hamilton, hamilton@mit.com
   \end{itemize}
   }}
   \caption{Added a speaker and sorted the list.}
   \label{fig:conf-add}
\end{subfigure}
\\~\\
\begin{subfigure}[b]{23em}\vspace{0pt}
   \fbox{\parbox{22.6em}{
   \textbf{PROGRAMMING CONFERENCE 2023}\\
   \textbf{Invited speakers}

   \begin{tabular}{| l | l | l |}
     \hline
     \textbf{Name} & \textbf{Email} & \textbf{Who} \\
     \hline \hline
     Adele Goldberg & adele@xerox.com & TP\\
     \hline
     Margaret Hamilton & hamilton@mit.com & JE\\
     \hline
     Betty Jean Jennings & betty@rand.com & JE\\
     \hline
   \end{tabular}
   \\~
   }}
   \caption{Refactored list into a table.}
   \label{fig:conf-tab}
\end{subfigure}
\hfill
\begin{subfigure}[b]{23em}\vspace{0pt}
   \fbox{\parbox{22.6em}{
   \textbf{PROGRAMMING CONFERENCE 2023}\\
   \textbf{Invited speakers}

   \begin{tabular}{| l | l | l |}
     \hline
     \textbf{Name} & \textbf{Email} & \textbf{Who} \\
     \hline \hline
     Ada Lovelace & ada@rsoc.ac.uk & \\
     \hline
     Adele Goldberg & adele@xerox.com & TP\\
     \hline
     Betty Jean Jennings & betty@rand.com & JE\\
     \hline
     Margaret Hamilton & hamilton@mit.com & JE\\
     \hline
   \end{tabular}
   }}
   \caption{Resulting document after change merging.}
   \label{fig:conf-fin}
\end{subfigure}
\end{adjustwidth}
\vspace{0.5em}
\caption{Conference organizer. Initial version of the document with two independent changes
and the final version of the document after the two changes are merged.}
\label{fig:conf}
\vspace{0.5em}
\end{figure}

\section{Conference Organizer: Merging edits in computational documents}
In this section, we consider challenges related to working with a document-based format akin to HTML.
We discuss two variants of the format. In the first format, a document
contains only data. In the second variant, the format is extended with computed values.
A computed value is a node in the document whose value is specified by a formula akin to those
used in spreadsheets. The equations can refer to other nodes in the document, possibly using
an absolute or a relative path and may involve a range of built-in functions, for example to
count the number of elements or to sum a sequence of numerical values.

We assume that the document can be
concurrently edited by multiple users in the same fashion as in local-first software, i.e.,
users may edit a local copy that then needs to be synchronized with changes made by other users.
The edits to the document can be done in a rich-text editor, or by using a more structure-aware
structure editor that can, for example, record changes as a sequence of high-level edit operations.

The concrete challenges are centered around the problem of collaborative conference planning
as illustrated in Figure~\ref{fig:conf}. A number of co-organizers are planning a conference
that will feature talks by several invited speakers. They need to agree on the speaker list,
contact speakers and calculate the conference budget.

\frameworkbox{
In this scenario, the only transient \emph{state} is the result of computed values. The content
of the document corresponds to the \emph{data} pace layer. A change in the data layer may invalidate
the computed value at the state layer. Formulas correspond to the \emph{code} layer and changing them,
again, invalidates the state. The \emph{schema} pace layer is implicit. A change to the document that
modifies content is a mere data change, but restructuring the document is an implicit schema
change. Such schema change that may affect the code pace layer.

The scenario assumes the \emph{decentralized} model. There are multiple program (document)
variants and it should be possible to propagate a change at any pace layer to other program
variants. The challenge may be solved using both the easier \emph{convergence} and the more
challenging \emph{divergence} model.}

\paragraph{Applications.}
One of the earliest systems to combine documents with computational elements
is Boxer \cite{diSessa86}. The challenges in this section apply to a wide range of
systems that can be seen as its successors. The specific scenario discussed here is
inspired by the use of Notion~\cite{Notion24}, which is a collaborative structured document
editing system. Treating documents as end-user programs has recently
been explored in Potluck~\cite{Litt2023}, which is local-only but includes formulas similar
to those in our challenge.

The challenges in this section are also relevant to work on computational scientific notebooks.
Jupyter \cite{Kluyver2016} uses a simple representation as a sequence of cells and does not have
a mechanism for embedding data in the document and referring to it, but numerous
project such as Nextjournal \cite{Nextjournal21} advance the format. The idea of producing
interactive, computational secientific documents is an active research area. Recent contributions
include Nota~\cite{Crichton2021} and Living Papers \cite{Heer2023}.

\challenge{Merging structured document edits}

First, the organizers need to agree on a list of speakers to invite, coordinate who contacts
whom and track the acceptance of invitations. They start with a document shown in
Figure~\ref{fig:conf-orig}. The challenge is to merge document edits that are done locally
and independently by two co-organizers. Specifically, consider the following two edits:

@@

\noindent
The system should be able to merge the changes and produce a final table shown in
Figure~\ref{fig:conf-fin}. The resulting table needs to include the additional speaker
created by the first organizer, use the order specified by the first organizer and use
the format defined by the second organizer. The newly added speaker should be reformatted
into the new format. For the newly added ``Organizer'' column, the second user may
specify default value or the newly added row may use an empty value.

\frameworkbox{
In this challenge, the change done by the first organizer is at the \emph{data} layer,
while the change done by the second organizer spans the \emph{schema} and \emph{data} layers.
In implementing the schema change, the second organizer transforms the affected document data.
The challenge is to apply the same transformation to the data independently added by the
second organizer.
}

\paragraph{Requirements.}
The challenge is concerned with merging two sequences edits to a base document.
A solution to the challenge should strive to satisfy the following goals:

\begin{itemize}
\item \emph{Completeness.} It should be possible to semi-automatically merge any two sequences
  of edits. In other words, the merging should always be defined, regardless of what edits the
  users perform, although user input may sometimes be necessary.
\item \emph{Commutativity.} The order of edits should not matter, regardless of their pace layer.
  If users $A$ and $B$ perform edits independently starting from the same initial document, the
  resulting document should be the same if the system treats edits from $A$ as occurring before
  the edits of $B$ and vice versa.
\item \emph{Conflict resolution.} It may not always be possible to merge conflicting edits.
  In this case, the system should interactively ask the user for guidance or, possibly, use
  default behavior specified by the user.
\item \emph{Structure editing.} The system may use a structure editor that provides high-level
  commands for operations such as reordering of elements or refactoring of a list into a table.
  In other words, we recognise that the challenge may not be solvable in a system based on plain
  text editing of document source code.
\end{itemize}

\paragraph{Remarks.}
As discussed earlier, this challenge does not make an explicit distinction between the schema
of the document and data. The implementing system may or may not maintain such distinction.
In our example, adding a new speaker and sorting the list is a mere change of the data, but the
refactoring of a list into a table is a schema change. The system may offer different user interface
for changes at different layers and leverage knowledge about such distinction, for example,
by handling the merging of schema-related and data-related changes differently.

\begin{figure}
\begin{adjustwidth}{-5.5em}{-5.5em}
\centering
\begin{subfigure}[b]{23em}\vspace{0pt}
   \fbox{\parbox{22.6em}{
   \textbf{PROGRAMMING CONFERENCE 2023}\\
   \textbf{Invited speakers}
   \begin{itemize}
    \item Adele Goldberg, adele@xerox.com
    \item Margaret Hamilton, hamilton@mit.com
    \item Betty Jean Jennings, betty@rand.com
   \end{itemize}
   \vspace{1.5em}
   \textbf{Conference budget}\\
   Travel cost per speaker:\\
   \hspace*{2em} \$1200 \\
   Number of speakers: \\
   \hspace*{2em} \texttt{=COUNT(/ul[id='speakers']/li)} \\
   Travel expenses:\\
   \hspace*{2em} \texttt{=/dl/dd[0] * /dl/dd[1]}
   }}
   \caption{Initial state of the document with formulas.}
   \label{fig:conf-budget}
\end{subfigure}
\hfill
\begin{subfigure}[b]{23em}\vspace{0pt}
   \fbox{\parbox{22.6em}{
   \textbf{PROGRAMMING CONFERENCE 2023}\\
   \textbf{Invited speakers}

   \begin{tabular}{| l | l | l |}
     \hline
     \textbf{Name} & \textbf{Email} & \textbf{Who} \\
     \hline \hline
     Ada Lovelace & ada@rsoc.ac.uk & \\
     \hline
     Adele Goldberg & adele@xerox.com & TP\\
     \hline
     Betty Jean Jennings & betty@rand.com & JE\\
     \hline
     Margaret Hamilton & hamilton@mit.com & JE\\
     \hline
   \end{tabular}

   \vspace{0.5em}
   \textbf{Conference budget}\\
   Travel cost per speaker:\\
   \hspace*{2em} \$1200 \\
   Number of speakers: \\
   \hspace*{2em} \texttt{=COUNT(/table[id='speakers']/tbody/tr)} \\
   Travel expenses:\\
   \hspace*{2em} \texttt{=/dl/dd[0] * /dl/dd[1]}
   }}\caption{Resulting document after schema change merging.}
   \label{fig:conf-finfin}
\end{subfigure}
\caption{Conference organizer. Formulas to calculate the budget have been added to the
original document (Figure~\ref{fig:conf-orig}) and those have to be merged with the
refactoring of a list into a table (Fig~\ref{fig:conf-tab}). Formulas in the final version
are updated to match with the new document structure.}
\label{fig:conf-calc}
\end{adjustwidth}
\end{figure}

\challenge{Applying schema edits in presence of code}

This challenge extends the previous problem. The conference organizers now also want to use
the document to manage the conference budget. In a very simplified form, one aspect
of this is calculating the estimated travel expenses for the speakers. For this,
they add a computed value to the document that counts the number of speakers
and multiplies the result by a fixed cost per speaker elsewhere in the document.

The scenario is shown in Figure~\ref{fig:conf-budget}. Assume two co-organizers start with
the same original document shown in Figure~\ref{fig:conf-orig}. One adds an additional
section with computed values, while the other performs the edits discussed earlier.
The challenge is, again, to merge the edits done independently by the co-organizers.

\begin{enumerate}
\item The first co-organizer adds the budget calculation as shown in Figure~\ref{fig:conf-budget}.
  This includes two formulas. The first selects all \texttt{li} elements of a \texttt{ul} element
  with ID \texttt{speakers} (not visible in the UI) and counts their number. The second formula
  multiplies the constant travel cost per speaker by the number of speakers. Here, we assume the
  new section uses the HTML definition list \texttt{dl} and the formula selects its first and
  second \texttt{dd} item, respectively.

\item The second co-organizer performs the refactoring and data edits discussed previously.
  They edit the speakers and refactor the structure to use a table, resulting in the document
  shown in Figure~\ref{fig:conf-fin}.
\end{enumerate}

\noindent
As before, the system should be able to merge the edits made to the document.

\frameworkbox{
In this challenge, the change done by the first organizer is at the \emph{code} layer, while
the change done by the second organizer is both at the \emph{data} layer and, implicitly,
at the \emph{schema} layer. A change solely at the \emph{data} layer would not affect code,
but a change at the \emph{schema} layer requires adapting some aspects of the code.
}

\paragraph{Remarks.}
Addressing the challenge requires that the refactoring of the document structure done in (1),
is also reflected in the code of the equations added in (2). As shown in
Figure~\ref{fig:conf-finfin}, in this specific case, the system needs to change the equation
\texttt{COUNT(/ul[id='speakers']/li)} to \texttt{COUNT(/table[id='speakers']/tbody/tr)}.
This is the case because the refactoring consists of four steps that each transform the
document and also need to update the code of the equation accordingly:

\begin{itemize}
    \item[--] Change the type of the \texttt{ul} element to \texttt{table}
    \item[--] Wrap the body of the \texttt{table} element with \texttt{tbody}
    \item[--] Change the type of all children of \texttt{tbody} from \texttt{li} to \texttt{tr}
    \item[--] Further split the body into two \texttt{td} elements
\end{itemize}

\noindent
It is expected that the system solving the challenge will need to be aware of those high-level
edits. In particular, they may be explicitly specified by the user (through an interactive
user interface for editing the document) or semi-automatically inferred.

\paragraph{Requirements.}
A solution to the challenge should meet the goals of the preceding challenge, i.e., completeness
(merging should be always defined), commutativity (order does not matter), conflict resolution
(ask for guidance if needed), convergence (everyone eventually wants the same document) and
structure editing (the system may capture high-level edits).

\subsection*{Challenge Extensions: Liveness and divergence}
So far, we assumed that the system follows the \emph{convergence} model, i.e., all variants
of the program eventually adopt all changes. A more difficult variant on the challenge is to also
support the \emph{divergence} model. In this case, some of the users continue using their variant
of the program without adopting all of the changes done by other users.

\frameworkbox{
In the \emph{divergence} model, it may be expected that users will want to adopt all changes
done at the \emph{data} pace layer. They may selectively choose some of the changes at the
\emph{code} and \emph{schema} layers. The challenge is maintaining the same data across multiple
program variants and merging changes done to the data based on different document schema.
}

The basic challenge outlined above focuses primarily on the local-first software scenario, but
it can be extended to also apply to the live programming scenario. In particular, when the
document contains computed values, those may be evaluated either explicitly by the user or
automatically on-the-fly. An edit can then invalidate some of the values (either by changing
the data that a computed value depends on, or by changing the equation used to compute the value).
A live programming system should be able to detect which computed values are affected by an edit
and, either invalidate those (requiring an explicit re-evaluation by the user) or automatically
re-evaluate them (and possibly highlight the affected values to inform the user).

\frameworkbox{
In the live extension, we also consider the transient \emph{state} pace layer. A system should
automatically detect when a change at the \emph{data} or \emph{code} layers affects the computed
value and forces it to be recomputed. Interestingly, a change at the \emph{schema} layer may not
always affect the \emph{state} layer, even if it schema change requires modifying formulas that
compute the state.
}
